---
title: "NHANES-Report_Week4_Exercise4"
author: "Ida M. de Oliveira"
date: "2026-10-04"
output: 
  pdf_document:
    toc: true
    highlight: tango
    latex_engine: xelatex
geometry: margin=1in
link-citations: yes
citecolor: red
urlcolor: blue
bibliography: references.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  fig.align = "center"
  )
pacman::p_load(rio, here, janitor, dplyr, readr, skimr, tidyr, knitr, kableExtra)
```

```{r create-bib, include=FALSE}
cat('
@article{nhanes2017,
  author       = {{Centers for Disease Control and Prevention (CDC)}},
  title        = {{National Health and Nutrition Examination Survey (NHANES) Data}},
  year         = {2013-2014; 2015-2016; 2017-2018},
  publisher    = {U.S. Department of Health and Human Services, Centers for Disease Control and Prevention},
  url          = {https://www.cdc.gov/nchs/nhanes/index.htm}
}
', file = "references.bib")
```
# Overview & Dataset Introduction

The datasets were obtained from the **National Health and Nutrition Examination Survey (NHANES)** *<https://wwwn.cdc.gov/nchs/nhanes/>*.
NHANES is a program of studies designed to assess the health and nutritional status of adults and children in the United States [@nhanes2017]. The resulting data are unique in that they combine interviews with physical examinations and laboratory data.
In this report, we present a descriptive examination of a cleaned, multi-wave NHANES dataset (**Demographic Data** and **Blood Pressure Data** (from Examination Data component)), across three NHANES survey cycles.

# Data Summary and Demographic Distribution

The dataset comprises $29,400$ total observations spanning three survey waves: $2015$ ($N = 10,175$), $2016$ ($N = 9,970$), and $2017$ ($N = 9,254$). 

Across all three waves, the gender proportion remains consistently balanced at approximately $49.1\%$ to $49.2\%$ male participants. Ethnic diversity is recorded using standard NHANES race/ethnicity categorizations, with White, Black, Mexican, Asian, Other and Other Hispanic populations represented.

```{r import}
# Import the cleaned dataset
nhanes_data <- import(here("data", "cleaned_NHANES.csv"))

# Compute overall size and male counts per wave
sample_info <- nhanes_data %>%
  group_by(wave_id) %>%
  summarise(
    Total_N = n(),
    Male_N = sum(gender == "Male", na.rm = TRUE),
    Male_Pct = (Male_N / Total_N) * 100,
    .groups = "drop"
  )

# Demographic breakdown by ethnicity_2
ethnicity_info <- nhanes_data %>%
  group_by(wave_id, ethnicity_2) %>%
  summarise(
    Count = n(),
    .groups = "drop"
  ) %>%
  left_join(sample_info %>% select(wave_id, Total_N), by = "wave_id") %>%
  mutate(
    Pct = (Count / Total_N) * 100,
    Formatted_Ethnicity = paste0(format(Count, big.mark = ","), 
                                 " (", sprintf("%.1f", Pct), "%)")
  ) %>%
  select(wave_id, ethnicity_2, Formatted_Ethnicity) %>%
  pivot_wider(names_from = ethnicity_2, values_from = Formatted_Ethnicity)

# Merge overall sample & gender info with ethnicity table
demo_table <- sample_info %>%
  mutate(
    `Total Sample Size (N)` = format(Total_N, big.mark = ","),
    `Male Count (%)` = paste0(format(Male_N, big.mark = ","), 
                              " (", sprintf("%.1f", Male_Pct), "%)")
  ) %>%
  select(wave_id, `Total Sample Size (N)`, `Male Count (%)`) %>%
  left_join(ethnicity_info, by = "wave_id") %>%
  rename(Wave = wave_id)

# Render formatted LaTeX table
demo_table %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "NHANES Sample Size, Gender, and Ethnicity Distribution Across Survey Waves",
    col.names = c("Wave", "Total N", "Male N (%)", 
                  "Asian", "Black", "Mexican", 
                  "Other", "Other Hispanic", "White"),
    align = "c"
    ) %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down")) %>%
  add_header_above(c(
    " " = 1, 
    "Sample Overview" = 2, 
    "Ethnicity Breakdown (N, %)" = 6
    )) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, bold = TRUE, color = "#1A365D") %>%
  row_spec(0, bold = TRUE) %>%
  footnote(general = "Percentages are calculated relative to the total wave sample size.")
```

Key highlights from the demographic summary in Table 1 include:

* **Sample Size**: Wave $2015$ contained the largest cohort ($N = 10,175$), followed by Wave $2016$ ($N = 9,970$) and Wave $2017$ ($N = 9,254$).
* **Sex Ratio**: The male representation was nearly identical across waves: $49.2\%$ in $2015$, $49.1\%$ in $2016$, and $49.2\%$ in $2017$.
* **Ethnic Composition**: White participants comprised the largest single ethnic group ($30.8\%$ to $36.1\%$), followed by  Black participants ($21.4\%$ to $22.9\%$), Mexican American ($14.8\%$ to $19.3\%$), Asian ($10.5\%$ to $12.6\%$), and Other Hispanic ($8.9\%$ to $13.1\%$).



# Systolic Blood Pressure Analysis

Systolic blood pressure (SBP) readings were measured up to four times during individual examination sessions. To establish a reliable metric, an **average systolic blood pressure** was calculated for each individual by averaging all non-missing SBP values across readings $1$ through $4$.


```{r sbp-table}
# Calculate average SBP per participant
nhanes_sbp <- nhanes_data %>%
  rowwise() %>%
  mutate(
    avg_sbp = mean(c_across(c(
      systolic_bp_1, systolic_bp_2, 
      systolic_bp_3, systolic_bp_4
      )), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  filter(!is.nan(avg_sbp))

# Compute summary statistics by wave
sbp_stats <- nhanes_sbp %>%
  group_by(wave_id) %>%
  summarise(
    N_Valid = n(),
    Min_SBP = min(avg_sbp),
    Max_SBP = max(avg_sbp),
    Median_SBP = median(avg_sbp),
    Mean_SBP = mean(avg_sbp),
    SD_SBP = sd(avg_sbp),
    .groups = "drop"
  ) %>%
  mutate(
    N_Valid = format(N_Valid, big.mark = ","),
    Min_SBP = sprintf("%.1f", Min_SBP),
    Max_SBP = sprintf("%.1f", Max_SBP),
    Median_SBP = sprintf("%.1f", Median_SBP),
    Mean_SBP = sprintf("%.1f", Mean_SBP),
    SD_SBP = sprintf("%.1f", SD_SBP)
  )

# Render formatted SBP table
sbp_stats %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Summary Statistics of Average Systolic Blood Pressure (mmHg) by Wave",
    col.names = c("Wave", "Valid N", "Min", "Max", "Median", "Mean", "Std Dev"),
    align = "c"
  ) %>%
  kable_styling(latex_options = c("striped"), 
                full_width = FALSE) %>%
  add_header_above(c(" " = 2, "Systolic Blood Pressure Metrics (mmHg)" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(6, bold = TRUE, color = "#2B6CB0", background = "#EDF2F7") %>%
  row_spec(0, bold = TRUE) %>%
  footnote(
    general = "Average SBP is calculated across all available individual readings. SBP values are measured in mmHg."
  )
```

As detailed in Table 2, valid blood pressure measurements were available for $21,604$ total participants ($7528$ in $2015$, $7363$ in $2016$, and $6713$ in $2017$). Analysis of average SBP reveals:

1. **Mean and Median Trends**: Mean average SBP rose slightly across survey waves, increasing from $118.3 \text{ mmHg}$ in Wave $2015$ to $120.4 \text{ mmHg}$ in Wave $2016$, and $121.7 \text{ mmHg}$ in Wave $2017$. Median SBP demonstrated a parallel trend ($115.3 \text{ mmHg}$, $117.3 \text{ mmHg}$, and $118.0 \text{ mmHg}$, respectively).
2. **Variability**: Standard deviations remained stable across $2015$ ($18.1 \text{ mmHg}$) and $2016$ ($18.4 \text{ mmHg}$), with a slight increase in $2017$ ($20.3 \text{ mmHg}$).
3. **Range**: Minimum recorded averages ranged from $64.7$ to $74.0 \text{ mmHg}$, while maximum recorded averages ranged from $228.7$ to $234.0 \text{ mmHg}$.

# References
